Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-08-31 papers/wdmerger_I/paper.tex

\author{Max P. Katz, Michael Zingale, Alan C. Calder, F. Douglas Swesty, Ann S. Almgren, Weiqun Zhang}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possibility that has 
merited recent theoretical attention is the white dwarf merger scenario.
This can naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
However, to date there have been few fully self-consistent simulations 
of binary white dwarf systems using mesh-based hydrodynamics, 
relative to other methods. This is the first paper in a series designed to 
describe simulations of these systems using a hydrodynamics code with 
the adaptive-mesh refinement approach. In this paper we describe our numerical 
methodology and discuss our implementation in the code \castro. \castro\ 
solves the Euler equations of hydrodynamics 
and the Poisson equation for self-gravity. Standard techniques for 
coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system, 
but recent advances in the literature have made advances on this 
problem and we discuss our implementation here. We also run an 
extensive set of test problems to determine whether our software can sufficiently
model a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems, and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

Type Ia supernovae (SNe Ia) are currently some of the most exciting
events to study in astrophysics. These bright, brief pulses of light
in the distant universe have led to a number of important discoveries
in recent years, including the discovery of the accelerated expansion
of the universe \citep{perlmutter1999,riess1998}. However, their
origin is shrouded in mystery. It has long been expected that these
events arise from the thermonuclear explosions of white dwarfs
\citep{hoyle_fowler:1960}, but the cause of these explosions is
uncertain. In particular, it is not clear what process causes the
temperatures in these white dwarfs to become hot enough for explosive
burning of their constituent nuclei. The model favored initially by the
community was the single-degenerate (SD) model
\citep{whelan_iben:1973}. Accretion of material from a companion star
such as a red giant would cause the star to approach the Chandrasekhar
mass, and in doing so the temperature and density in the center would
become sufficient for thermonuclear fusion to proceed. However, in
recent years many alternative progenitor models have been discussed. A
leading candidate for explaining some or most of these explosions is
the double-degenerate (DD) model, in which two white dwarfs merge and
the merged object reaches the conditions necessary for a thermonuclear
ignition \citep{ibentutukov:1984,webbink:1984}. Another is the double
detonation scenario, where accretion of material onto a
sub-Chandrasekhar mass white dwarf leads to a detonation inside the
accreted envelope, sending a compressional wave into the
core of the star that triggers a secondary detonation. A recent
review of the progenitor models can be found in
\citet{hillebrandt:2013}.

@@

The first attempts to model the results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star---a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate. \cite{tutukov_yungelson:1979}
observed that this could happen if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

@@

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} pointed out that the previously mentioned simulations 
generally shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional grid-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of looking at the long-term dynamical evolution of binary
white dwarf systems after constructing exact equilibrium initial
conditions. Earlier work placed the stars too close together 
and ignored the effects of tidal forces that change the shape of the 
secondary, leading to the merger
happening artificially too quickly. When the initial conditions are
constructed in exact equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. However, one limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

@@

Given the above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. Most of the existing results that study 
the viability of double degenerate systems as progenitors for
Type Ia supernovae (that is, including a realistic 
equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as excellent conservation of
angular momentum. However, the question of whether a prompt detonation
in a merger actually happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a grid-based code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. However, there are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason for caution is that other than the
most recent results of \cite{kashyap:2015}, no simulation has self-consistently
resulted in a thermonuclear detonation. Reproducibility of the detonation 
through direct numerical simulation is critical for building 
confidence in this progenitor model. \MarginPar{Direct numerical
  simulation has a specific meaning to CFD community.  That is the
  simulation resolve the viscous scale.  So maybe we should remove the
word 'direct'.}

This is the first in a series of papers designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:implementation} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to demonstrate that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

We use the unsplit piecewise-parabolic method (PPM) solver in \castro\
to advance the hydrodynamics system in time \citep{ppmunsplit}.  A
number of changes were made to the solver, which are detailed in the
Appendix.  These changes bring the algorithm more in line with that of
\cite{ppm}. \castro\ as originally released featured a slightly modified
version of the higher resolution limiters of
\cite{colella_sekora:2008}, which can be accessed in the code using
\texttt{castro.ppm\_type = 2}. The advantage of this limiter is that
it preserves physical extrema rather than clipping them off as in the
original approach of \cite{ppm}. However, we found these limiters to
be unsatisfactory for our problem. There are many regions in our
problem with large density gradients (such as the interface between
the star's atmosphere and the ambient gas outside of it) and in these
regions the algorithm can yield negative densities. This often results
from the limiters interpreting these gradients as being true
minima. As a result, we use the original limiter, which is strictly
monotonicity preserving in the parabolic profiles it generates; this
is activated with \texttt{castro.ppm\_type = 1}.

@@

The equation of state (EOS) for our simulations is the Helmholtz EOS
\citep{timmes_swesty:2000}. This models an electron-positron gas of
arbitrary relativity and degeneracy over a wide range of temperatures
and densities. Thermodynamic quantities are calculated as derivatives
of the Helmholtz free energy, and the values are interpolated from a
table. The natural variables of the Helmholtz free energy are
temperature and density, and calling the EOS is simplest in this
form. However, in hydrodynamics we often have the density, and
internal energy as independent variables, and we want to obtain the
temperature, pressure, and other quantities. To do this, we employ a
Newton-Raphson iteration over the temperature (given some sufficient
starting guess) until we find the temperature that corresponds to the
desired internal energy. Sometimes this process fails to converge and
the iterative value approaches zero. In these cases we employ a
``floor'' that limits how low the temperature can go (typically 
$10^5$ K). There is a choice here how to proceed: we can either
assign this floor value to the temperature and let that zone be
thermodynamically inconsistent (the original behavior in \castro), or
we can adjust the internal energy to be thermodynamically consistent
with the temperature, at the cost of violating energy conservation. We
have found in some test problems of strong one-dimensional shocks that reach 
the temperature floor that the latter yields more accurate results. 
However, allowing the equation of state call to update the 
internal energy can actually result in significant changes to the 
total energy of the system over long periods of time, 
due not just to resets in low-density zones but also to small 
inconsistencies between the energy given to the EOS and the energy 
that is consistent with the returned temperature. These inconsistencies
are dependent on the threshold we set that determines how close the 
returned energy should be to the original input energy. While this 
threshold is typically very small in an individual zone (a relative 
difference of $10^{-8}$ by default in \castro), over time and given 
a large number of zones, this can result in a significant energy 
drift. This is a serious enough problem that we opt for the energy 
conserving approach for our simulations.

@@

The effect of gravity on the hydrodynamical evolution is typically
incorporated by the use of a source term for the momentum and energy
equations. In a finite volume methodology, the momentum source term 
as traditionally written appears in integral form as
\begin{equation}
  \left.\frac{\partial (\rho \mathbf{u})}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{g}\, dV
\end{equation}
and for the energy source term the analogue it is
\begin{equation}
  \left.\frac{\partial (\rho E)}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV \label{eq:cell_center_gravity_source}.
\end{equation}
Here $\Delta V$ is the cell's volume.
In most hydrodynamics codes these are discretized as $\rho\,
\mathbf{g}$ and $\rho\, \mathbf{u}\,\cdot\mathbf{g}$, respectively, 
where $\rho$, $\mathbf{u}$, and $\mathbf{g}$ 
are evaluated at the zone center.

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
However, we believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

In a system with self-gravity, total momentum is conserved if the spatial domain
includes all of the mass of the system. This must be the 
case because each mass element exerts an equal and opposite gravitational force 
on every other mass element. However, the standard approach does not necessarily
guarantee that momentum is conserved numerically. We cannot represent a vacuum state 
in our code, so there is a small but non-zero density on the edge of the grid. 
This allows momentum to leak out of the domain even if the gravitational source term 
is written in an explicitly conservative manner. To see this, one can write the 
density in terms of the potential and then consider its spatial discretization. For simplicity,
we will consider one spatial dimension.
\begin{align}
  -\rho_{i}  \frac{d\Phi_{i}}{dx} &= -\frac{1}{4\pi G} \frac{d^2\Phi_i}{dx^2} \frac{d \Phi}{dx} \notag \\
  &= -\frac{1}{4\pi G} \left[\frac{\Phi_{i-1} - 2 \Phi_{i} + \Phi_{i+1}}{\Delta x^2}\right] \left[ \frac{\Phi_{i+1} - \Phi_{i-1}}{2\Delta x} \right] \notag \\
  &= -\frac{1}{8\pi G \Delta x^3} \left[ \Phi_{i+1}^2 - \Phi_{i-1}^2 - 2\Phi_i\left(\Phi_{i+1} - \Phi_{i-1}\right) \right] \label{eq:momentum_discretization}
\end{align}
It is easy to verify that adding the source terms for the current zone and the two zones 
to the left and right will result in complete cancellation of the source terms.
However, if the potential if non-zero outside of the domain, then there will be
momentum lost or gained from the grid, which will be encapsulated in the ghost cells
just outside the domain. In addition, when we replace the Laplacian above by the full
three-dimensional stencil including the $y$ and $z$ derivatives, depending on the
discretization these will not be cancelled at all. This latter problem can be resolved by
writing the momentum update in an explicitly conservative way.

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term 
for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form.
The flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum in the domain interior stays constant to
within numerical roundoff error. This result can be derived by analytically recasting 
\autoref{eq:momentum_discretization}. In the continuum limit, the two momentum
formulations are identical. Thus the latter has been advocated by, for example, 
\cite{jiang:2013} for the ATHENA code. However, in a finite discretization 
the divergence of the gravitational acceleration is no longer guaranteed to equal
the zone density. In particular, we find that the mixing of the gravitational accleration components
means that the truncation error in the gravitational field can lead to large errors
that imply a density much different than the zone's actual density. This is especially
problematic in a simulation with a low-density ambient medium, where even a small error 
in the momentum update can lead to large changes in a zone's momentum. By continuing to explicitly
use the cell density in the momentum update, we can avoid this possibility: the size of the update
will always be suitably small if the zone's density is small. Thus for our simulations
we continue to use the standard source term for the momentum.

@@

Analytical solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = -\sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{eq:direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions, so we have implemented it as an option in \castro).

>>>>

\author{Max P. Katz, Michael Zingale, Alan C. Calder, F. Douglas Swesty, Ann S. Almgren, Weiqun Zhang}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possibility that has 
merited recent theoretical attention is the white dwarf merger scenario.
This can naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been few fully self-consistent simulations 
of binary white dwarf systems using mesh-based hydrodynamics, 
relative to other methods. This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
the adaptive-mesh refinement approach. In this paper we describe our numerical 
methodology and discuss our implementation in the code \castro. \castro\ 
solves the Euler equations of hydrodynamics 
and the Poisson equation for self-gravity. Standard techniques for 
coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system, 
but recent advances in the literature allow progress on this 
problem and we discuss our implementation here. We present an 
extensive set of test problems demonstrating the extent to which our software can sufficiently
model a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

Type Ia supernovae (SNe Ia) are among the most exciting
events to study in astrophysics. These bright, brief pulses of light
in the distant universe have led to a number of important discoveries
in recent years, including the discovery of the accelerated expansion
of the universe \citep{perlmutter1999,riess1998}. Their origin, though,
is shrouded in mystery. It has long been expected that these
events arise from the thermonuclear explosions of white dwarfs
\citep{hoyle_fowler:1960}, but the cause of these explosions is
uncertain. In particular, it is not clear what process causes the
temperatures in these white dwarfs to become hot enough for explosive
burning of their constituent nuclei. The model favored initially by the
community was the single-degenerate (SD) model
\citep{whelan_iben:1973}. Accretion of material from a companion star
such as a red giant would cause the star to approach the Chandrasekhar
mass, and in doing so the temperature and density in the center would
become sufficient for thermonuclear fusion to proceed. In
recent years the focus has shifted to a number of alternative progenitor models. A
leading candidate for explaining some or most of these explosions is
the double-degenerate (DD) model, in which two white dwarfs merge and
the merged object reaches the conditions necessary for a thermonuclear
ignition \citep{ibentutukov:1984,webbink:1984}. Another is the double
detonation scenario, where accretion of material onto a
sub-Chandrasekhar mass white dwarf leads to a detonation inside the
accreted envelope, sending a compressional wave into the
core of the star that triggers a secondary detonation. A recent
review of the progenitor models can be found in
\citet{hillebrandt:2013}.

@@

The first attempts to model the results of the merger process came in the
1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon
ignition would occur in the more massive white dwarf as it accreted
mass near the Eddington rate from the less massive white dwarf
overflowing its Roche lobe. \cite{saionomoto:1985} tracked the
evolution of the flame and found that it propagated quiescently into
the center, converting the carbon-oxygen white dwarf into an
oxygen-neon-magnesium white dwarf. This would then be followed by
collapse into a neutron star---a result with significantly different
observational properties compared to a SN Ia. This scenario, termed
accretion-induced collapse, would be avoided only if the accretion
rate were well below the Eddington rate (see, e.g., \cite{fryer:1999}
for a discussion of the possible implications of the accretion-induced 
collapse scenario). \cite{tutukov_yungelson:1979}
observed that the collapse could be avoided if the mass loss from the secondary
was higher than the Eddington rate and thus the accreted material
formed an accretion disc, which might rain down on the primary more
slowly. The main finding was that double degenerate systems would not
obviously lead to Type Ia supernovae.

@@

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} pointed out that the previously mentioned simulations 
generally shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional grid-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of looking at the long-term dynamical evolution of binary
white dwarf systems after constructing equilibrium initial
conditions. Earlier work placed the stars too close together 
and ignored the effects of tidal forces that change the shape of the 
secondary, leading to the merger
happening artificially too quickly \citep{fryer:2008}. When the initial conditions are
constructed in exact equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. One limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

@@

Given the considerable research into the double degenerate problem 
described above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. Most of the existing results that study 
the viability of double degenerate systems as progenitors for
Type Ia supernovae (that is, including a realistic 
equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as excellent conservation of
angular momentum. A drawback relates to the fact that whether a prompt detonation
in a merger happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a grid-based code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. However, there are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason for caution is that other than the
most recent results of \cite{kashyap:2015}, no simulation has self-consistently
resulted in a thermonuclear detonation. Reproducibility of the detonation 
through numerical simulation is critical for building 
confidence in this progenitor model.

This paper is the first in a series designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:implementation} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to verify that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

We use the unsplit piecewise-parabolic method (PPM) solver in \castro\
to advance the hydrodynamics system in time \citep{ppmunsplit}.  A
number of changes were made to the solver, which are detailed in the
Appendix.  These changes bring the algorithm more in line with that of
\cite{ppm}. \castro\ as originally released featured a slightly modified
version of the higher resolution limiters of
\cite{colella_sekora:2008}, which can be accessed in the code using
\texttt{castro.ppm\_type = 2}. The advantage of this limiter is that
it preserves physical extrema rather than clipping them off as in the
original approach of \cite{ppm}. Despite the advantages of this limiter 
we have found it to be unsatisfactory for our problem. There are many regions in our
problem with large density gradients (such as the interface between
the star's atmosphere and the ambient gas outside of it) and in these
regions the algorithm can yield negative densities. This often results
from the limiters interpreting these gradients as being true
minima. As a result, we use the original limiter, which is strictly
monotonicity preserving in the parabolic profiles it generates; this
is activated with \texttt{castro.ppm\_type = 1}.

@@

The equation of state (EOS) for our simulations is the Helmholtz EOS
\citep{timmes_swesty:2000}. This models an electron-positron gas of
arbitrary relativity and degeneracy over a wide range of temperatures
and densities. Thermodynamic quantities are calculated as derivatives
of the Helmholtz free energy, and the values are interpolated from a
table. The natural variables of the Helmholtz free energy are
temperature and density, and calling the EOS is simplest in this
form. In hydrodynamics we often have the density and
internal energy as independent variables, and we want to obtain the
temperature, pressure, and other quantities. To do this, we employ a
Newton-Raphson iteration over the temperature (given some sufficient
starting guess) until we find the temperature that corresponds to the
desired internal energy. Sometimes this process fails to converge and
the iterative value approaches zero. In these cases we employ a
``floor'' that limits how low the temperature can go (typically 
$10^5$ K). There is a choice here how to proceed: we can either
assign this floor value to the temperature and let that zone be
thermodynamically inconsistent (the original behavior in \castro), or
we can adjust the internal energy to be thermodynamically consistent
with the temperature, at the cost of violating energy conservation. We
have found in some test problems of strong one-dimensional shocks that reach 
the temperature floor that the latter yields more accurate results. 
However, allowing the equation of state call to update the 
internal energy can actually result in significant changes to the 
total energy of the system over long periods of time, 
due not just to resets in low-density zones but also to small 
inconsistencies between the energy given to the EOS and the energy 
that is consistent with the returned temperature. These inconsistencies
are dependent on the threshold we set that determines how close the 
returned energy should be to the original input energy. While this 
threshold is typically very small in an individual zone (a relative 
difference of $10^{-8}$ by default in \castro), over time and given 
a large number of zones, this can result in a significant energy 
drift. This is a serious enough problem that we opt for the energy 
conserving approach for our simulations.

@@

The effect of gravity on the hydrodynamical evolution is typically
incorporated by the use of a source term for the momentum and energy
equations. In a finite volume methodology, the momentum source term 
as traditionally written appears in integral form as
\begin{equation}
  \left.\frac{\partial (\rho \mathbf{u})}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{g}\, dV
\end{equation}
and for the energy source term it is
\begin{equation}
  \left.\frac{\partial (\rho E)}{\partial t}\right|_{\text{grav}} = \frac{1}{\Delta V} \int \rho \mathbf{u}\cdot\mathbf{g}\, dV \label{eq:cell_center_gravity_source}.
\end{equation}
Here $\Delta V$ is the cell's volume.
In most hydrodynamics codes these are discretized as $\rho\,
\mathbf{g}$ and $\rho\, \mathbf{u}\,\cdot\mathbf{g}$, respectively, 
where $\rho$, $\mathbf{u}$, and $\mathbf{g}$ 
are evaluated at the zone center.

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
We believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

In a system with self-gravity, total momentum is conserved if the spatial domain
includes all of the mass of the system. This must be the 
case because each mass element exerts an equal and opposite gravitational force 
on every other mass element. However, the standard approach does not necessarily
guarantee that momentum is conserved numerically. We cannot represent a vacuum state 
in our code, so there is a small but non-zero density on the edge of the grid. 
This allows momentum to leak out of the domain even if the gravitational source term 
is written in an explicitly conservative manner. To see this, one can use the Poisson equation to write the 
density in terms of the potential and then consider its spatial discretization. For simplicity,
we will consider one spatial dimension and a uniform discretization. Analogous results 
may be readily obtained for the non-uniform case.
\begin{align}
  -\rho_{i}  \frac{d\Phi_{i}}{dx} &= -\frac{1}{4\pi G} \frac{d^2\Phi_i}{dx^2} \frac{d \Phi_i}{dx} \notag \\
  &= -\frac{1}{4\pi G} \left[\frac{\Phi_{i-1} - 2 \Phi_{i} + \Phi_{i+1}}{\Delta x^2}\right] \left[ \frac{\Phi_{i+1} - \Phi_{i-1}}{2\Delta x} \right] \notag \\
  &= -\frac{1}{8\pi G \Delta x^3} \left[ \Phi_{i+1}^2 - \Phi_{i-1}^2 - 2\Phi_i\left(\Phi_{i+1} - \Phi_{i-1}\right) \right] \label{eq:momentum_discretization}
\end{align}
It is easy to verify that adding the source terms for the current zone and the two zones 
to the left and right will result in complete cancellation of the source terms.
The catch is that if the potential if non-zero outside of the domain, then there will be
momentum lost or gained from the grid, which will be encapsulated in the ghost cells
just outside the domain. In addition, when we replace the Laplacian above by the full
three-dimensional stencil including the $y$ and $z$ derivatives, depending on the
discretization these may not be cancelled at all. This latter problem can be resolved by
writing the momentum update in an explicitly conservative way.

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term 
for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form.
The flux at any zone boundary is added to one cell and
subtracted from another, so that the total momentum in the domain interior stays constant to
within numerical roundoff error. This result can be derived by analytically recasting 
\autoref{eq:momentum_discretization}. In the continuum limit, the two momentum
formulations are identical. Thus the latter has been advocated by, for example, 
\cite{jiang:2013} for the ATHENA code. A significant limitation to this approach is that in a finite discretization 
the divergence of the gravitational acceleration is no longer guaranteed to equal
the zone density. In particular, we find that the mixing of the gravitational accleration components
means that the truncation error in the gravitational field can lead to large errors
that imply a density much different than the zone's actual density. This is especially
problematic in a simulation with a low-density ambient medium, where even a small error 
in the momentum update can lead to large changes in a zone's momentum. By continuing to explicitly
use the cell density in the momentum update, we can avoid this possibility: the size of the update
will always be suitably small if the zone's density is small. Thus for our simulations
we continue to use the standard source term for the momentum.

@@

Analytical solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. On a finite computational domain, however, it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, the boundary values can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = -\sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{eq:direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions, so we have implemented it as an option in \castro).

In a typical simulation we place the boundaries of the domain far
enough away from the region containing most of the mass that some
method of approximation to this direct summation is justified. Many
approaches exist in the literature. The original release of \castro\
featured the crudest possible approximation: a monopole prescription,
where the boundary values were computed by summing up all the mass on
the domain and treating it as a point source at the domain
center. This is correct only for a spherically symmetric mass
distribution, and therefore is best suited for problems like
single-star Type Ia supernova simulations (e.g.$\ $\cite{malone:2014})
that employ self-gravity. For a problem like that of a binary star system
with significant departures from spherical symmetry, this assumption
fails to produce accurate boundary values, which we find in \autoref{sec:kepler}
results in a significant drift of the center of the mass of the system over time.
